Map{R}educe in {MPI} for Large-scale Graph Algorithms

### 2010-05-04 paper/graph.tex
More edits.
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@605 3fc257bb-8550-0410-81b3-a5c103da83f7

We begin with a MapReduce procedure for creating large, sparse,
randomized graphs, since they are the input for the algorithms
discussed below.  R-MAT graphs ~\cite{RMAT} are recursively generated
graphs with power-law degree distributions.  They are commonly used to
represent web and social networks.  The user specifices six parameters
that define the graph: the number of vertices $N$ and edges $M$, and
four parameters $a$, $b$, $c$, $d$ that sum to 1.0 and are discussed
below.  The algorithm in Figure \ref{fig:rmat} generates $M$ unique
non-zero entries in a sparse $N \times N$ matrix $A$, where each entry
$A_{ij}$ represents an edge between graph vertices $(V_i,V_j)$.

@@

By using more than one MapReduce object, we can improve the performance
of the R-MAT generation algorithm.  In the improved algorithm
(Figure~\ref{fig:rmat2}), we emit newly created edges into MapReduce object
$E_{new}$ and {\it aggregate()} the edges to processors.  
We then add those edges
to the MapReduce object $E_{old}$, containing all previously generated
unique edges; the edges of $E_{old}$ are already aggregated to processors
using the same mapping as $E_{new}$.  We can then use the {\it compress()}
function to reduce duplicates locally on each processor.
Performance improvements
in this algorithm result from the reduced number of KV pairs communicated in
{\it aggregate()},
compared to the collate operation in the original algorithm, 
particularly in later iterations of the generation where few
edges are created.

@@

 \caption{Improved MapReduce algorithm for R-MAT graph generation.}

@@

The algorithm in Figure \ref{fig:pr} performs these
iterations.  The nonzeros of matrix $A$ are initialized as described above;
the PageRank vector $x$ is initialized uniformly.  A MapReduce object $MT$
contains the indices of all-zero rows of $A$, corresponding to vertices
with no outlinks.  In step 1 of a PageRank interation, $MT$ is used 
in dot products to compute adjustments to the
PageRank vector to represent uniform probability of going from a leaf page
to any other page.  Adjustments for random jumps from any page are also
computed in this step.  Step 2, the matrix-vector product $A^T x$, is the most
expensive part of the computation, requiring two passes over the nonzero
structure of $A$.  In the first pass, {\it collate()} 
gathers all row entries $a_{ij}$ with their
associated $x_i$ entry, and {\it reduce()} computes $a_{ij} x_i$.  
In the second pass, 
{\it collate()} gathers, for each $j$, all contributions to the
column sum $\sum_i a_{ij} x_i$, which is computed by a second {\it reduce()}. 
Steps 3 and 4 adjust and scale the 
product vector; the work is proportional to $|V|$.  Step 6 computes the
residual to determine whether the PageRank vector has converged; work is
again proportional to $|V|$.  Throughout the algorithm, we exploit the
ability to call MPI\_Allreduce to compute global norms and residuals.

@@

The entire PageRank algorithm can be made more efficient by calling
{\it aggregate()} on the MapReduce objects representing the matrix $A$, 
the empty matrix rows $MT$, and the vector $x$ once before the PageRank 
iterations begin.  Since these three MapReduce
objects have the same key types, pre-aggregating the objects moves all
Key-Value pairs with a given key to a single processor.  Thus, many of
the PageRank operations become local operations, replacing 
the {\it collate()} operation (which is equivalent to {\it aggregate()} 
followed by {\it convert()}) with only a local {\it convert()} call. 
Figure~\ref{fig:pr2} shows the
improved algorithm; modified operations are marked with an asterisk.
This strategy of pre-aggregating to improve data locality is useful in
many MapReduce algorithms, but assumes that keys are always mapped to
the same processor.

@@

Input: \\
$*$\> Pre-aggregated MapReduce object $A$ containing $|V| \times |V|$ matrix:  Key = $i$, Value = $[j, a_{ij}]$ \\
$*$\> Pre-aggregated MapReduce object $MT$ containing indices of empty rows of $A$:  Key = $i$, Value = 0. \\
$*$\> Pre-aggregated MapReduce object $x$ containing initial PageRank vector:  Key = $i$, Value = $1/|V|$. \\
While (${residual} > {tolerance}$) \\
\> Compute contribution for random jumps and zero-outdegree vertices. \\
\> \> Add: $x$ to $MT$. \\
$*$\> \> Convert $MT$:  Vertex is Key. \\
\> \> Reduce $MT$:   \\
\> \> \> Input:  Key = $i$; MultiValue = $x_i, 0$ if $i \in {MT}$; MultiValue = $x_i$ if $i \notin {MT}$. \\
\> \> \> Compute:  Local contributions $c_{local}$. \\
\> \> \> Output:  If ${nvalues} = 2$, Key = $i$; Value = 0. \\
\> \> MPI\_Allreduce $c_{local}$ with MPI\_SUM to compute global contribution $c_{global}$. \\
\> Compute $y = A^T x$.  \\
\> \> Copy $x$ to $y$. \\
\> \> Add $A$ to $y$. \\
$*$\> \> Convert $y$:  Key is row index of $A$. \\
\> \> Reduce $y$: \\
\> \> \>  Input: Key = $i$; Multivalue = $i, [j, a_{ij}]$ for nonzeros $a_{ij}$ in $A$ \\
\> \> \>  Output:  Key = $j$; Value = $a_{ij} y_i$. \\
\> \> Collate $y$:  Key is column index of $A$. \\
\> \> Reduce $y$:  \\
\> \> \> Input:  Key = $j$; Multivalue = $a_{ij} y_i$ for all non-zero entries $a_{ij}$ of column $j$ \\
\> \> \> Output:  Key = $j$; Value = $\sum_i a_{ij} y_i$. \\
\> Add contribution $c_{global}$ to $y$; compute local max norm $n_{local}$ of $y$. \\
\> \> Map $y$:  \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Compute: $y_i = y_i + c_{global}$; if $y_i > {n_{local}}$, $n_{local} = y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i$. \\
\> \> MPI\_Allreduce $n_{local}$ with MPI\_MAX to compute global max norm $n_{global}$. \\
\> Scale $y$ by $n_{global}$. \\
\> \> Map $y$: \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i / {n_{global}}$. \\
\> Compute residual. \\
\> \> Add $y$ to $x$. \\
$*$\> \> Convert $x$:  Key is index $i$. \\
\> \> Reduce $x$:   \\
\> \> \> Input:  Key = $i$; Multivalue = $x_i, y_i$. \\
\> \> \> Compute:  if $|x_i - y_i| > {resid}$, ${resid} = |x_i - y_i|$ \\
\> \> MPI\_Allreduce with MPI\_MAX to compute $residual$. \\
\> $x = y$; \\
Output:  PageRank vector $x$.
  \end{tabbing}
 \end{minipage}}\end{center}

 \caption{Improved MapReduce algorithm for PageRank vertex ranking; pre-aggregating the MapReduce objects allows many communication intensive {\it collate()} 
operations to be replaced by with strictly local {\it convert()} operations (as marked by the asterisks).}

@@

A connected component of a graph is a set of vertices where all pairs
of vertices in the set are connected by a path of edges.  A sparse
graph may contain many such components.  Figure \ref{fig:cc} outlines
a MapReduce algorithm that labels each vertex in a graph with a
component ID.  All vertices in the same component are labelled with
the same ID, which is the ID of a vertex in the component.  We assume
an input graph of undirected edges $(V_i,V_j)$.  This exposition also
follows the connected-component algorithm presented in \cite{Cohen},
with the addition of logic that load-balances data across processors
when one or a few giant components exist in the graph.

@@

The algorithm begins (before the iteration loop) by assigning each
vertex to its own component or ``zone'', so that $Z_i$ = $V_i$.  Each
iteration will grow the zones, one layer of neighbors at a time.  As
zones collide due to shared edges, a winner is chosen (the smaller
zone ID), and vertices in the losing zone are reassigned to the
winning zone.  When the iterations complete, each zone will have
become a fully connected component.  The algorithm thus finds all
connected components in the graph simultaneously.  The number of
iterations required depends on the largest diameter of any component
in the graph.

@@

Since each edge was emitted twice, the second {\it collate()}
operation collects the zone assignments for its two vertices together.
If the two zone IDs are different, the second {\it reduce()} operation
chooses a winner (the min of the 2 IDs), and emits the loser ID as a
key, with the winning ID as a value.  If no zone ID changes are
emitted, the algorithm is finished, and the iteration exits.

The third {\it map()} operation inverts the vertex/zone key/value
pairs to become zone/vertex pairs.  The third {\it add()} operation
adds the changing zone assignments to the set of key/value pairs.  The
third collate() can then collects all the vertices of a zone and zero
or more reassignments for the zone ID.  Since a zone could collide
with multiple other zones on the same iteration due to shared edges,
the new zone ID becomes the minimum ID of any of the neighboring
zones.  If no zone reassignment value appears in the multi-value, the
zone ID is unchanged.  The final reduce() a key/value pair for each
vertex in the zone, with the vertex as a key and the new zone ID as a
value.

Note that if a graph has only a few components, then the third {\it
collate()} operation, which keys on the zone ID, may generate a few
very large key/multi-value (KMV) pairs.  For example, if the graph is
fully connected, then on the last iteration, a single KMV pair will
contain all vertices in the graph and be assigned to one processor.
This imbalance in memory and computational work can lead to poor
parallel performance of the overall algorithm.  To counter this
effect, the various operations of stage 3 include extra logic.  The
idea is to partition zones whose vertex count exceeds a user-defined
threshhold into $P$ sub-zones, where $P$ is the number of processors.
The 64-bit integer that stores the zone ID also stores a bit flag
indicating the zone has been partitioned and a set of bits that encode
the processor ID.

During the third {\it map()} operation, if the zone has been
partitioned, then the vertex is assigned to a random processor and the
processor ID bits are added to the zone ID, as indicated by the
$Z_i^+$ notation in Figure \ref{fig:cc}.  Likewise, if $Z_i$ has been
partitioned, the third {\it add()} operation emits the zone ID change
key/value pair ($Z_i$,$Z_{winner}$) as ($Z_i^+$,$Z_{winner}$).  In
this case ($Z_i$,$Z_{winner}$), is emitted not once, but $P+1$ times,
once for each processor, and once as if $Z_i$ had not been partitioned
(for a reason discussed below).

@@

Note that this logic does not guarantee that the partition bits of the
zone IDs for all the vertices in a single zone will be set
consistently on a given iteration.  For example, an unpartitioned zone
with a small ID may consume a partitioned zone.  The vertices from the
partitioned zone will retain their partitioned status, but the
original vertices in the small zone may not set the partition bit of
their zone IDs.  On subsequent iterations, the third {\it add()}
operation emits $P+1$ copies of new zone reassignments for both
partitioned and unpartioned zone IDs, to insure all vertices in the
zone will know the reassignment information.

@@

Map: assign random values to each vertex \\
Clone: convert edge key/values directly to key/multivalues \\
Create: empty MapReduce MRv for maximal independent set vertices \\
\\
while $N_{edges} > 0$: \\
\> 1 Reduce: \> Determine WINNER/LOSER vertex of each edge \\
       \> \> \> input Key = $E_{ij}$ with $V_i < V_j$, Multivalue = NULL or LOSER \\
       \> \> \> if either value is LOSER, emit nothing \\
       \> \> \> $V_w$ = WINNER vertex, $V_l$ = LOSER vertex \\
       \> \> \> output Key = $V_w$, Value = $V_l$ WINNER \\
       \> \> \> output Key = $V_l$, Value = $V_w$ LOSER \\
\> 1 Collate: \> Vertex as key \\
\> 2 Reduce: \> Find WINNER vertices \\
       \> \> \> input Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> \> if all $V$ are WINNERs: \\
      \> \> \> \> for each $V$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $Vi$ LOSER \\
       \> \> \> else: \\
      \> \> \> \> for each $V$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $V_i$ \\
\> 2 Collate: \> Vertex as key \\
\> 3 Reduce: \> Find LOSER vertices \\
       \> \> \> input Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> \> if any $V$ is LOSER: \\
      \> \> \> \> for each $V$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $Vi$ LOSER \\
       \> \> \> else: \\
      \> \> \> \> for each $V$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $V_i$ \\
\> 3 Collate: \> Vertex as key \\
\> 4 Reduce: \> Emit WINNER vertices and LOSER edges \\
       \> \> \> input Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> \> if all $V$ are LOSERs: \\
      \> \> \> \> output to MRv, Key = $V_i$, Value = NULL \\
       \> \> \> for each $V$ in Multivalue: \\
      \> \> \> \> if $V$ is LOSER: \\
     \> \> \> \> \> output Key = $E_{ij}$ with $V_i < V_j$, Value = LOSER \\
      \> \> \> \> else: \\
     \> \> \> \> \> output Key = $E_{ij}$ with $V_i < V_j$, Value = NULL \\
\> 4 Collate: \> Edge as key \\

@@

Before the iterations begin, a random value is assigned to each
vertex, which will be used to compare pairs of vertices.  The vertex
ID itself could be used as the random value, but since the vertices
eventually selected for the MIS depend on the randomization, this may
introduce an unwanted bias.  Instead a random number generator can be
used to assign a consistent random value $R_i$ to each vertex in each
edge (via a map() operation), which is then carried along with the
vertex ID through each stage of the algorithm.  In the notation of
Figure \ref{fig:luby}, each $V_i$ is then really two quantities, a
vertex ID (1 to $N$) and $R_i$.  Alternatively, the $R_i$ can be used
to relabel the vertices in a random manner, assigning each a new,
unique vertex ID from 1 to $N$.  In this case, $V_i$ is simply the new
vertex ID, and the vertices in the final MIS could be remapped to
recover their original IDs.  These operations could be performed with
a few extra MapReduce steps, outside the iteration loop.

The pre-iteration clone() operation converts the initial list of edge
key/values one by one to key/multivalue pairs, with the edge as the
key, and NULL as the value.  This is so the iterations can begin with
a reduce() operation, without need for a collate() operation.  A
second empty MapReduce object is also initialized, which will
accumulate MIS vertices as the iterations proceed.

At each iteration, the current set of edges are examined to identify
winning vertices.  Vertices adjacent to winners are flagged as losers,
and all edges adjacent to losers are flagged for removal at the start
of the next iteration.  Vertices with all losing edges are flagged as
winners.  At each iteration edges are removed from the graph, and
winning vertices are added to the MIS.  When the graph is empty, the
MIS is complete.

The first reduce() operation flags the two vertices in each edge as a
potential winner or loser.  This is done by comparing the two random
values for the vertices (or the randomized vertex IDs as explained
above).  The first collate() operation thus produces a multivalue for
each vertex $V_i$ which contains all the vertices it shares an edge
with, and associated winner/loser flags.  In the second reduce()
operation, if $V_i$ won the comparison with all its edge neighbors,
then it is a ``winner'' vertex.  All vertices connected to the winner
are emitted with a loser flag.  If $V_i$ is not an overall winner, its
edge vertices are emitted without the loser flag.  The second
collate() operation collects this new information for each vertex.

In the third reduce() operation, if $V_i$ was flagged as a loser by
any edge neighbor who was a winner, then it becomes a losing vertex
and emits all its edge vertices with a loser flag.  Otherwise it emits
all its edge vertices without the flag.

In the fourth reduce() operation, some vertices will have all their
edge vertices flagged as losers.  These vertices are all winners,
either the ones identified in the second reduce() operation, or
additional vertices who would become singletons (having no edges) when
edges containing a loser vertex are removed (on the next iteration).
All of these winning vertices are emitted to the MapReduce object that
stores the accumulating MIS.

All edges of each vertex are also emitted, retaining the loser flag if
they have it.  They are emitted with $E_ij$ as their key, with $V_i <
V_j$.  This is to insure both copies of the edge come together via the
fourth collate() operation.  The multivalue for each edge will now
have two values, either of which may be NULL or a loser flag.

@@

As with PageRank and R-MAT generation, 
significant savings in communication and execution time
can be achieved by pre-aggregating certain MapReduce objects and 
storing data in more than one MapReduce object.  The improved algorithm
is shown in Figure~\ref{fig:sssp2}.  The MapReduce objects storing
vertices and edges are aggregated to processors only once at the beginning
of the computation.  Then, in each iteration of 
SSSP, only a small number of new candidate distances are generated.
The improved algorithm uses MapReduce object $U$ to store these updates.
Only the updates are communicated; each update is aggregated to the 
same processor that stores the corresponding vertex and its edges.
Communication is performed using $U$, so the total amount of communication
needed is small compared to the original algorithm (which aggregates all
graph edges in each iteration).  Updated vertex distances are stored back
in the the MapReduce object for vertices.

@@

 \caption{Improved MapReduce algorithm for single-source shortest path (SSSP)
 determination.  Pre-aggregating the vertices and edges, and later aggregating only updates to vertex distances reduces communication and execution time.}

>>>>

We begin with a MapReduce procedure for creating large, sparse,
randomized graphs, since they are the input for the algorithms
discussed below.  R-MAT graphs ~\cite{RMAT} are recursively generated
graphs with power-law degree distributions.  They are commonly used to
represent web and social networks.  The user specifices six parameters
that define the graph: the number of vertices $N$ and edges $M$, and
four parameters $a$, $b$, $c$, $d$ that sum to 1.0 and are discussed
below.  The algorithm in Figure~\ref{fig:rmat} generates $M$ unique
non-zero entries in a sparse $N \times N$ matrix $A$, where each entry
$A_{ij}$ represents an edge between graph vertices $(V_i,V_j)$.

@@

By using more than one MapReduce object, we can improve the performance
of the R-MAT generation algorithm.  In the enhanced algorithm
(Figure~\ref{fig:rmat2}), we emit newly created edges into MapReduce object
$E_{new}$ and {\it aggregate()} the edges to processors.  
We then add those edges
to the MapReduce object $E_{old}$, containing all previously generated
unique edges; the edges of $E_{old}$ are already aggregated to processors
using the same mapping as $E_{new}$.  We can then use the {\it compress()}
function to reduce duplicates locally on each processor.
Performance improvements
in this algorithm result from the reduced number of KV pairs communicated in
{\it aggregate()},
compared to the collate operation in the original algorithm, 
particularly in later iterations of the generation where few
edges are created.

@@

 \caption{Enhanced MapReduce algorithm for R-MAT graph generation.}

@@

The algorithm in Figure~\ref{fig:pr} performs these
iterations.  The nonzeros of matrix $A$ are initialized as described above;
the PageRank vector $x$ is initialized uniformly.  A MapReduce object $MT$
contains the indices of all-zero rows of $A$, corresponding to vertices
with no outlinks.  In step 1 of a PageRank interation, $MT$ is used 
in dot products to compute adjustments to the
PageRank vector to represent uniform probability of going from a leaf page
to any other page.  Adjustments for random jumps from any page are also
computed in this step.  Step 2, the matrix-vector product $A^T x$, is the most
expensive part of the computation, requiring two passes over the nonzero
structure of $A$.  In the first pass, {\it collate()} 
gathers all row entries $a_{ij}$ with their
associated $x_i$ entry, and {\it reduce()} computes $a_{ij} x_i$.  
In the second pass, 
{\it collate()} gathers, for each $j$, all contributions to the
column sum $\sum_i a_{ij} x_i$, which is computed by a second {\it reduce()}. 
Steps 3 and 4 adjust and scale the 
product vector; the work is proportional to $|V|$.  Step 6 computes the
residual to determine whether the PageRank vector has converged; work is
again proportional to $|V|$.  Throughout the algorithm, we exploit the
ability to call MPI\_Allreduce to compute global norms and residuals.

@@

The entire PageRank algorithm can be made more efficient by calling
{\it aggregate()} on the MapReduce objects representing the matrix $A$, 
the all-zero matrix rows $MT$, and the vector $x$ once before the PageRank 
iterations begin.  Since these three MapReduce
objects have the same key types, pre-aggregating the objects moves all
Key-Value pairs with a given key to a single processor.  Thus, many of
the PageRank operations become local operations, replacing 
the {\it collate()} operation (which is equivalent to {\it aggregate()} 
followed by {\it convert()}) with only a local {\it convert()} call. 
Figure~\ref{fig:pr2} shows the
enhanced algorithm; modified operations are marked with an asterisk.
This strategy of pre-aggregating to improve data locality is useful in
many MapReduce algorithms, but assumes that keys are always mapped to
the same processor.

@@

Input: \\
$*$\> Pre-aggregated MapReduce object $A$ containing $|V| \times |V|$ matrix:  Key = $i$, Value = $[j, a_{ij}]$ \\
$*$\> Pre-aggregated MapReduce object $MT$ containing indices of all-zero rows of $A$:  Key = $i$, Value = 0. \\
$*$\> Pre-aggregated MapReduce object $x$ containing initial PageRank vector:  Key = $i$, Value = $1/|V|$. \\
While (${residual} > {tolerance}$) \\
\> Compute contribution for random jumps and zero-outdegree vertices. \\
\> \> Add: $x$ to $MT$. \\
$*$\> \> Convert $MT$:  Vertex is Key. \\
\> \> Reduce $MT$:   \\
\> \> \> Input:  Key = $i$; MultiValue = $x_i, 0$ if $i \in {MT}$; MultiValue = $x_i$ if $i \notin {MT}$. \\
\> \> \> Compute:  Local contributions $c_{local}$. \\
\> \> \> Output:  If ${nvalues} = 2$, Key = $i$; Value = 0. \\
\> \> MPI\_Allreduce $c_{local}$ with MPI\_SUM to compute global contribution $c_{global}$. \\
\> Compute $y = A^T x$.  \\
\> \> Copy $x$ to $y$. \\
\> \> Add $A$ to $y$. \\
$*$\> \> Convert $y$:  Key is row index of $A$. \\
\> \> Reduce $y$: \\
\> \> \>  Input: Key = $i$; Multivalue = $i, [j, a_{ij}]$ for nonzeros $a_{ij}$ in $A$ \\
\> \> \>  Output:  Key = $j$; Value = $a_{ij} y_i$. \\
\> \> Collate $y$:  Key is column index of $A$. \\
\> \> Reduce $y$:  \\
\> \> \> Input:  Key = $j$; Multivalue = $a_{ij} y_i$ for all non-zero entries $a_{ij}$ of column $j$ \\
\> \> \> Output:  Key = $j$; Value = $\sum_i a_{ij} y_i$. \\
\> Add contribution $c_{global}$ to $y$; compute local max norm $n_{local}$ of $y$. \\
\> \> Map $y$:  \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Compute: $y_i = y_i + c_{global}$; if $y_i > {n_{local}}$, $n_{local} = y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i$. \\
\> \> MPI\_Allreduce $n_{local}$ with MPI\_MAX to compute global max norm $n_{global}$. \\
\> Scale $y$ by $n_{global}$. \\
\> \> Map $y$: \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i / {n_{global}}$. \\
\> Compute residual. \\
\> \> Add $y$ to $x$. \\
$*$\> \> Convert $x$:  Key is index $i$. \\
\> \> Reduce $x$:   \\
\> \> \> Input:  Key = $i$; Multivalue = $x_i, y_i$. \\
\> \> \> Compute:  if $|x_i - y_i| > {resid}$, ${resid} = |x_i - y_i|$ \\
\> \> MPI\_Allreduce with MPI\_MAX to compute $residual$. \\
\> $x = y$; \\
Output:  PageRank vector $x$.
  \end{tabbing}
 \end{minipage}}\end{center}

 \caption{Enhanced MapReduce algorithm for PageRank vertex ranking; pre-aggregating the MapReduce objects allows many communication intensive {\it collate()} 
operations to be replaced by with strictly local {\it convert()} operations (as marked by the asterisks).}

@@

A connected component of a graph is a set of vertices where all pairs
of vertices in the set are connected by a path of edges.  A sparse
graph may contain many such components.  Figure~\ref{fig:cc} outlines
a MapReduce algorithm that labels each vertex in a graph with a
component ID.  All vertices in the same component are labelled with
the same ID, which is the ID of a vertex in the component.  We assume
an input graph of undirected edges $(V_i,V_j)$.  This exposition also
follows the connected-component algorithm presented in \cite{Cohen},
with the addition of logic that load-balances data across processors
when one or a few giant components exist in the graph.

@@

The algorithm begins (before the iteration loop) by assigning each
vertex to its own component or ``zone,'' so that $Z_i$ = $V_i$.  Each
iteration will grow the zones, one layer of neighbors at a time.  As
zones collide due to shared edges, a winner is chosen (the smaller
zone ID), and vertices in the losing zone are reassigned to the
winning zone.  When the iterations complete, each zone will have
become a fully connected component.  The algorithm thus finds all
connected components in the graph simultaneously.  The number of
iterations required depends on the largest diameter of any component
in the graph.

@@

Since each edge was emitted twice, the second {\it collate()}
operation collects the zone assignments for its two vertices together.
If the two zone IDs are different, the second {\it reduce()} operation
chooses a winner (the smaller of the two IDs), and emits the loser ID as a
key, with the winning ID as a value.  If no zone ID changes are
emitted, the algorithm is finished, and the iteration exits.

The third {\it map()} operation inverts the vertex/zone key/value
pairs to become zone/vertex pairs.  The third {\it add()} operation
adds the changing zone assignments to the set of key/value pairs.  The
third collate() can then collect all the vertices of a zone and zero
or more reassignments for the zone ID.  Since a zone could collide
with multiple other zones on the same iteration due to shared edges,
the new zone ID becomes the minimum ID of any of the neighboring
zones.  If no zone reassignment value appears in the multi-value, the
zone ID is unchanged.  The final reduce() a key/value pair for each
vertex in the zone, with the vertex as a key and the new zone ID as a
value.

Note that if a graph has only a few components, the third {\it
collate()} operation, which collates by the zone ID, may generate a few
very large key/multi-value (KMV) pairs.  For example, if the graph is
fully connected, then on the last iteration, a single KMV pair will
contain all vertices in the graph and be assigned to one processor.
This imbalance in memory and computational work can lead to poor
parallel performance of the overall algorithm.  To counter this
effect, the various operations of stage 3 include extra logic.  The
idea is to partition zones whose vertex count exceeds a user-defined
threshhold into $P$ sub-zones, where $P$ is the number of processors.
The 64-bit integer that stores the zone ID also stores a bit flag
indicating the zone has been partitioned and a set of bits that encode
the processor ID.

During the third {\it map()} operation, if the zone has been
partitioned, then the vertex is assigned to a random processor and the
processor ID bits are added to the zone ID, as indicated by the
$Z_i^+$ notation in Figure~\ref{fig:cc}.  Likewise, if $Z_i$ has been
partitioned, the third {\it add()} operation emits the zone ID change
key/value pair ($Z_i$,$Z_{winner}$) as ($Z_i^+$,$Z_{winner}$).  In
this case ($Z_i$,$Z_{winner}$), is emitted not once, but $P+1$ times,
once for each processor, and once as if $Z_i$ had not been partitioned
(for a reason discussed below).

@@

Note that this logic does not guarantee that the partition bits of the
zone IDs for all the vertices in a single zone will be set
consistently on a given iteration.  For example, an unpartitioned zone
with a small ID may consume a partitioned zone.  The vertices from the
partitioned zone will retain their partitioned status, but the
original vertices in the small zone may not set the partition bit of
their zone IDs.  On subsequent iterations, the third {\it add()}
operation emits $P+1$ copies of new zone reassignments for both
partitioned and unpartioned zone IDs, to ensure all vertices in the
zone will know the reassignment information.

@@

Map: assign random values to each vertex \\
Clone: convert edge key/values directly to key/multivalues \\
Create: empty MapReduce object MRv for maximal independent set vertices \\
\\
while $N_{edges} > 0$: \\
\> 1 Reduce: \> Determine WINNER/LOSER vertex of each edge \\
       \> \> \> input Key = $E_{ij}$ with $V_i < V_j$, Multivalue = NULL or LOSER \\
       \> \> \> if either value is LOSER, emit nothing \\
       \> \> \> $V_w$ = WINNER vertex, $V_l$ = LOSER vertex \\
       \> \> \> output Key = $V_w$, Value = $V_l$ WINNER \\
       \> \> \> output Key = $V_l$, Value = $V_w$ LOSER \\
\> 1 Collate: \> Vertex as key \\
\> 2 Reduce: \> Find WINNER vertices \\
       \> \> \> input Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> \> if all $V$ are WINNERs: \\
      \> \> \> \> for each $V_j$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $V_i$ LOSER \\
       \> \> \> else: \\
      \> \> \> \> for each $V_j$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $V_i$ \\
\> 2 Collate: \> Vertex as key \\
\> 3 Reduce: \> Find LOSER vertices \\
       \> \> \> input Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> \> if any $V$ is LOSER: \\
      \> \> \> \> for each $V_j$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $V_i$ LOSER \\
       \> \> \> else: \\
      \> \> \> \> for each $V_j$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $V_i$ \\
\> 3 Collate: \> Vertex as key \\
\> 4 Reduce: \> Emit WINNER vertices and LOSER edges \\
       \> \> \> input Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> \> if all $V$ are LOSERs: \\
      \> \> \> \> output to MRv, Key = $V_i$, Value = NULL \\
       \> \> \> for each $V_j$ in Multivalue: \\
      \> \> \> \> if $V_j$ is LOSER: \\
     \> \> \> \> \> output Key = $E_{ij}$ with $V_i < V_j$, Value = LOSER \\
      \> \> \> \> else: \\
     \> \> \> \> \> output Key = $E_{ij}$ with $V_i < V_j$, Value = NULL \\
\> 4 Collate: \> Edge as key \\

@@

Before the iterations begin, a random value is assigned to each
vertex, which will be used to compare pairs of vertices.  The vertex
ID itself could be used as the random value, but since the vertices
eventually selected for the MIS depend on the randomization, this may
introduce an unwanted bias.  Instead a random number generator can be
used to assign a consistent random value $R_i$ to each vertex in each
edge (via a {\it map()} operation), which is then carried along with the
vertex ID through each stage of the algorithm.  In the notation of
Figure~\ref{fig:luby}, each $V_i$ is then really two quantities, a
vertex ID (1 to $N$) and $R_i$.  Alternatively, the $R_i$ can be used
to relabel the vertices in a random manner, assigning each a new,
unique vertex ID from 1 to $N$.  In this case, $V_i$ is simply the new
vertex ID, and the vertices in the final MIS could be remapped to
recover their original IDs.  These operations could be performed with
a few extra MapReduce steps, outside the iteration loop.

The pre-iteration {\it clone()} operation converts the initial list of edge
key/values one by one to key/multivalue pairs, with the edge as the
key, and NULL as the value.  Thus, the iterations can begin with
a {\it reduce()} operation, without need for a {\it collate()} operation.  A
second empty MapReduce object is also initialized, which will
accumulate MIS vertices as the iterations proceed.

At each iteration, the current set of edges is examined to identify
winning vertices.  Vertices adjacent to winners are flagged as losers,
and all edges adjacent to losers are flagged for removal at the start
of the next iteration.  Vertices with all losing edges are flagged as
winners.  At each iteration edges are removed from the graph, and
winning vertices are added to the MIS.  When the graph is empty, the
MIS is complete.

The first {\it reduce()} operation flags the two vertices in each edge as a
potential winner or loser.  This is done by comparing the two random
values for the vertices (or the randomized vertex IDs as explained
above).  The first {\it collate()} operation thus produces a multivalue for
each vertex $V_i$ which contains all the vertices it shares an edge
with, and associated winner/loser flags.  In the second {\it reduce()}
operation, if $V_i$ won the comparison with all its edge neighbors,
then it is a ``winner'' vertex.  All vertices connected to the winner
are emitted with a loser flag.  If $V_i$ is not an overall winner, its
edge vertices are emitted without the loser flag.  The second
{\it collate()} operation collects this new information for each vertex.

In the third {\it reduce()} operation, if $V_i$ was flagged as a loser by
any edge neighbor who was a winner, then it becomes a losing vertex
and emits all its edge vertices with a loser flag.  Otherwise it emits
all its edge vertices without the flag.

In the fourth {\it reduce()} operation, some vertices will have all their
edge vertices flagged as losers.  These vertices are all winners,
either the ones identified in the second {\it reduce()} operation, or
additional vertices who would become singletons (having no edges) when
edges containing a loser vertex are removed (on the next iteration).
All of these winning vertices are emitted to the MapReduce object that
stores the accumulating MIS.

All edges of each vertex are also emitted, retaining the loser flag if
they have it.  They are emitted with $E_{ij}$ as their key, with $V_i <
V_j$.  This is to ensure both copies of the edge come together via the
fourth {\it collate()} operation.  The multivalue for each edge will now
have two values, either of which may be NULL or a loser flag.

@@

As with PageRank and R-MAT generation, 
significant savings in communication and execution time
can be achieved by pre-aggregating certain MapReduce objects and 
storing data in more than one MapReduce object.  The enhanced algorithm
is shown in Figure~\ref{fig:sssp2}.  The MapReduce objects storing
vertices and edges are aggregated to processors only once at the beginning
of the computation.  Then, in each iteration of 
SSSP, only a small number of new candidate distances are generated.
The enhanced algorithm uses MapReduce object $U$ to store these updates.
Only the updates are communicated; each update is aggregated to the 
same processor that stores the corresponding vertex and its edges.
Communication is performed using $U$, so the total amount of communication
needed is small compared to the original algorithm (which aggregates all
graph edges in each iteration).  Updated vertex distances are stored back
in the the MapReduce object for vertices.

@@

 \caption{Enhanced MapReduce algorithm for single-source shortest path (SSSP)
 determination.  Pre-aggregating the vertices and edges, and later aggregating only updates to vertex distances reduces communication and execution time.}
